---
title: "NYPD Shooting Incident Data Report"
date: "11/7/2021"
output: pdf_document
---

```{r lib, include=FALSE}
library(tidyverse)
library(lubridate)
```

## Data Gathering
The data in this analysis is from <https://catalog.data.gov/dataset/nypd-shooting-incident-data-historic> that contains records of shooting incidents from 2006-2020 in New York.

In this analysis, I will focus on four fields namely INCIDENT_KEY, BORO, OCCUR_TIME, and OCCUR_DATE. I would then convert the OCCUR_DATE to date format and call it "DATE" as well as OCCUR TIME to time format and call it "TIME".
```{r Select_data}

url_in <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
nypd_data <- read.csv(url_in)

nypd_by_date <- nypd_data %>%
  mutate(TIME = hour(hms(nypd_data$OCCUR_TIME)),
         DATE = mdy(nypd_data$OCCUR_DATE)) %>%
  select(INCIDENT_KEY, BORO, DATE, TIME)

head(nypd_by_date)
summary(nypd_by_date)
```
## Data Analysis

I would like to start my analysis on looking at the shooting cases by year.\
Data below shows shooting cases trend over the years and shows that cases started to decline starting in 2013 until 2019 but interestingly started a dramatic increase by 2020.

```{r year, message = FALSE}
nypd_by_date %>% 
  group_by(DATE = year(DATE)) %>%
  summarise(CASES = n()) %>% 
  ggplot(aes(x = DATE, y = CASES)) +
  geom_line(color = "blue")
```

Another interesting thing to look is how the cases behave by BORO. Data below shows shooting cases trend over the years by BORO and shows that the trend is uniform for across all BOROs. See below chart.

```{r year_by_boro, message = FALSE}
nypd_by_date %>% 
  group_by(DATE = year(DATE), BORO) %>%
  summarise(CASES = n()) %>% 
  ggplot(aes(x = DATE, y = CASES)) +
  geom_line(aes(color = BORO))
```

Now, I specifically wanted to look at the time when these events occur and possibly look at a pattern. The chart below shows total cases of shooting by each time interval. Note that in this assumption, the day will start at 5AM and so the time column had a little transformation to make 12 AM to 4 AM greater than 11 PM.

```{r by_time, message = FALSE}
nypd_by_time <- nypd_by_date %>%
  group_by(TIME) %>%
  summarise(AVG.CASES = mean(n()))

nypd_by_time$TIME <- as.numeric(nypd_by_time$TIME)

nypd_by_time$TIME2 <- ifelse(nypd_by_time$TIME == 0 , 24 ,
                             ifelse(nypd_by_time$TIME == 1 , 25 , 
                                    ifelse(nypd_by_time$TIME == 2 , 26 , 
                                           ifelse(nypd_by_time$TIME == 3 , 27 , 
                                                  ifelse(nypd_by_time$TIME == 4 , 28, nypd_by_time$TIME)))))
nypd_by_time %>%
  ggplot()+
  geom_point(aes(x = TIME2, y = AVG.CASES)) +
  xlab("TIME 5AM - 4AM")+
  theme(axis.text.x=element_blank())
```



```{r pred, message = FALSE}
lm = lm(AVG.CASES ~ TIME2, data = nypd_by_time)  

nypd_by_time$PRED = predict(lm)

nypd_by_time %>%
  ggplot()+
  geom_point(aes(x = TIME2, y = AVG.CASES)) +
  geom_line(aes(x = TIME2, y = PRED), color = "red") +
  xlab("TIME 5AM - 4AM")+
  theme(axis.text.x=element_blank())
```

## Conclusion and Bias


The cases has the same trend by BORO and that there was a significant increase in year 2020. I notice that generally the cases tend to go up as the day goes later or as the day approaches nighttime and that the surrounding is dark. This was confirmed by a simple linear model on the 2nd chart. Obviously, there is a strong correlation between time of day and the cases of shooting.

There are a lot of factors that is not considered in this simple analysis such as the population density and diversity, income levels, location, and so on.  On the latest census data for example, BRONX has almost 40% less population than QUEENS but shows that it has more cases of shooting, and so population alone cannot be a determining factor of cases of shooting. 
